---
title: "Dimensionality reduction: PCA, NMF, tSNE and UMAP"
output: html_document
date: "2025-12-04"
---

```{r,message=FALSE,warning=FALSE}
library(factoextra)
library(RColorBrewer)
```

## The data

Let us consider a single-cell RNA sequencing (scRNA-seq) dataset.
The dataset consist of 500 single cells and expression measurements of 6148 different genes.
The expression matrix is already normalized.
```{r}
#Load the data
load("/Users/dsesia/Desktop/Teaching/Clustering/Data/normalized_matrix.subset.RData")
load("/Users/dsesia/Desktop/Teaching/Clustering/Data/cell_types.subset.RData")
```

We also have information about the cell type of each cell. In particular there are 7 different types of cells:
```{r}
table(cell_types$CellType)
```


## PCA

Now, let us perform dimensionality reduction using principal component analysis.
In particular we start by performing PCA using the R function prcomp().

```{r}
pca=prcomp(data,scale = FALSE)
summ=summary(pca)
```

Let us see how much variance of the data is explained by the first 10 principal components.
```{r,out.width="40%"}
fviz_eig(pca)
```

Let us visualize the PCA results by plotting the observations in the two dimensional space of the 2 first principal components.

```{r,out.width="50%"}
# assign each observation a color based on their cell types
temp <- as.numeric(as.factor(cell_types$CellType))
colsamples <- brewer.pal(7, "Paired")[temp]

# plot the data
plot(pca$x[,1:2], main="PCA",col=colsamples,pch = 19,xlab = paste0("PC1 (",100*round(summ$importance[2,1],3),"%)"),ylab = paste0("PC2 (",100*round(summ$importance[2,2],3),"%)"))
legend(x="bottomleft",legend=levels(as.factor(cell_types$CellType)),
     fill=brewer.pal(7, "Paired"),cex=0.8)
```


## Non-negative Matrix Factorization (NMF)

NMF R package paper: [Gaujoux, R., Seoighe, C. A flexible R package for nonnegative matrix factorization. BMC Bioinformatics 11, 367 (2010).](https://doi.org/10.1186/1471-2105-11-367)

```{r,message=FALSE,warning=FALSE}
#install.packages("NMF")
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("Biobase")
library(NMF)
```

```{r}
# For simplicity (running time) we consider just the 500 most variable genes
var=apply(data,2,function(x) var(x,na.rm = TRUE))
var=var[order(var,decreasing = TRUE)]
more.var=names(var)[1:500]
data.var=data[,more.var]
```

We run NMF using the R function nmf(). 
The parameter "rank" is the factorization rank (i.e the number of columns in matrix W, that is the number of dimensions that we want) 

*Recall*: **X ≈ WH**

```{r}
# run NMF
r.nmf=nmf(data.var,rank=2)

# exctract the W and H matrices from the nmf run result
nmf.w <- NMF::basis(r.nmf)
nmf.h <- NMF::coef(r.nmf)
```

We now visualize the data in a 2 dimensional space.

```{r,out.width="50%"}
# dataset for plot
nmf_df <- as.data.frame(nmf.w)
n.dim=ncol(nmf_df)
colnames(nmf_df) <- paste0("dim",1:n.dim)
nmf_df$celltype <- factor(cell_types$CellType) # add cell type info

# create the scatter plot
ggplot(nmf_df, aes(x=dim1, y=dim2, color=celltype)) + geom_point() + theme_light() +
  scale_colour_manual(values = brewer.pal(7, "Paired"))
```

**Basis components matrix (W)**:

"Metagene" expression matrix. Dim = n_samples x rank 

The samples are clustered (hierarchical clustering) based on the metagenes. 
The second colored column on the left represents the metagene that most contribute to the expression profile of each sample.

```{r,out.width="50%"}
# Data frame for sample annotation(for heatmap)
ann_color=list(brewer.pal(3, "RdYlBu")[-2],brewer.pal(7, "Paired")) 
basismap(r.nmf,annRow=factor(cell_types$CellType),annColors=ann_color)
```

**Mixture coefficient matrix (H)**:

"Metagene" matrix. Dim = rank x n_genes

It show the contribution of the genes to each metagene. 

```{r,out.width="50%"}
ann_color=list(brewer.pal(3, "RdYlBu")[-2]) 
coefmap(r.nmf,annColors=ann_color)
```

**How to choose the rank?**

As suggested in:

* [J. Brunet, P. Tamayo, T.R. Golub, J.P. Mesirov, Metagenes and molecular pattern discovery using matrix factorization, Proc. Natl. Acad. Sci. U.S.A., 101 (12) 4164-4169](https://doi.org/10.1073/pnas.0308531101) 
* [Gaujoux, R., Seoighe, C. A flexible R package for nonnegative matrix factorization. BMC Bioinformatics 11, 367 (2010).](https://doi.org/10.1186/1471-2105-11-367): 

A common way of deciding on r is to try different values, compute some **quality measure** of the results, and choose the best value according to this quality criteria. 

For example, one of the most common approach is to use the **cophenetic correlation coefficient**. 

Steps: 

  1) Performing NMF for multiple candidate ranks, with multiple runs for each rank to ensure stability (remember that due to random initialization of W and H, results might vary from one run to the other).
  2) Evaluating quality metrics such as the cophenetic correlation coefficient (CCC).

The CCC measures the stability of clustering across runs for a given choice of r.
(In this context, observations are assigned to clusters based on their dominant basis component. So if r=2 there would be 2 clusters, and each observation is assigned to one of the two clusters based on the dominant component). 

A higher CCC value indicates greater consistency and reproducibility of clustering results for a given rank.


Brunet et al. suggested choosing the smallest value of r for which CCC starts decreasing.


If you want more info on how the **cophenetic correlation coefficient** is computed, check the **"Model Selection"** section of [Brunet et al.](https://doi.org/10.1073/pnas.0308531101). 

```{r}
# Run NMF for multiple candidate ranks (2,3,4,...,10), with multiple runs (nrun=10) for each rank to ensure stability:
#estim.r <- nmf(data.var, 2:10, nrun=10, seed=123456)

# here I just saved the result of the above computation, since it takes a while! 
#save(estim.r,file="/Users/dsesia/Desktop/Teaching/Dimensionality reduction/nmf_choose_r_2_10.RData")
load(file="/Users/dsesia/Desktop/Teaching/Dimensionality reduction/nmf_choose_r_2_10.RData")
```

```{r}
# Look at the cophenetic correlation coefficient:
plot(estim.r$measures[,c("rank","cophenetic")],type = "b",main="Cophenetic correlation coefficient")
```

In this example, we can choose rank=3, as it is the smallest rank for which the cophenetic correlation coefficient is high before decreasing, providing a balance between stability and simplicity. For rank=4, we observe a decrease, indicating reduced stability. 
(If you want to look for deeper insight, in this case, it might be worth it to also explore r=5 since it also shows high stability. Is k=5 capturing additional underlying structure in the data or is it just adding complexity?)

NMF with factorization **rank = 3**

```{r}
# run NMF
r.nmf.3=nmf(data.var,rank=3)

# exctract the W and H matrices from the nmf run result
nmf.w <- NMF::basis(r.nmf.3)
nmf.h <- NMF::coef(r.nmf.3)
```

We now visualize the data in a 2 dimensional space. (Space of the first two NMF factors.)

```{r,out.width="50%"}
# dataset for plot
nmf_df <- as.data.frame(nmf.w)
n.dim=ncol(nmf_df)
colnames(nmf_df) <- paste0("dim",1:n.dim)
nmf_df$celltype <- factor(cell_types$CellType) # add cell type info

# create the scatter plot
ggplot(nmf_df, aes(x=dim1, y=dim2, color=celltype)) + geom_point() + theme_light() +
  scale_colour_manual(values = brewer.pal(7, "Paired"))
```

**Basis components matrix (W)**:

```{r,out.width="50%"}
ann_color=list(brewer.pal(3, "RdYlBu"),brewer.pal(7, "Paired"))
basismap(r.nmf.3,annRow=factor(cell_types$CellType),annColors=ann_color)
```

**Mixture coefficient matrix (H)**:

```{r,out.width="50%"}
ann_color=list(brewer.pal(3, "RdYlBu")) 
coefmap(r.nmf.3,annColors=ann_color)
```


Useful resources for NMF in R: 

* [NMF Vignette - introduction](https://cran.r-project.org/web/packages/NMF/vignettes/NMF-vignette.pdf)
* [NMF Vignette - heatmaps](https://cran.r-project.org/web/packages/NMF/vignettes/heatmaps.pdf)


Let us now consider **non-linear** methods like t-SNE and umap.

## t-SNE

Let us now vizualize our data by using t-SNE.
```{r}
# install.packages("Rtsne")
library(Rtsne)
```

In particular, we perform t-SNE using the first 20 principal components that we previously computed. This step allows us to remove noise while preserving the major sources of variation in the data, which helps t-SNE run faster and produce more stable, meaningful embeddings.

```{r,out.width="50%"}
data.pc=pca$x[,1:20]
tsne=Rtsne(data.pc,pca=FALSE)
plot(tsne$Y, main="tsne",col=colsamples,pch = 19,xlab = "tsne1",ylab = "tsne2")
legend(x="bottomleft",legend=levels(as.factor(cell_types$CellType)),
     pch=19,col=brewer.pal(7, "Paired"),cex=0.8)
```

Remark: type "help(Rtsne)" for more info on the possible input parameters of the function. \
The parameter "perplexity" can be interpreted as a guess about the number of close neighbors each point has. Default perplexity is 30.  
The perplexity parameter reflect the focus given to local and global aspects of your data (if the perplexity is small, more attention is given to local variations, higher values focus more on broader patterns).

Try, for example, perplexity=2. (--> many small “micro-clusters”)

```{r,out.width="50%"}
tsne=Rtsne(data.pc,pca=FALSE,perplexity = 2)
plot(tsne$Y, main="tsne",col=colsamples,pch = 19,xlab = "tsne1",ylab = "tsne2")
legend(x="bottomleft",legend=levels(as.factor(cell_types$CellType)),
     pch=19,col=brewer.pal(7, "Paired"),cex=0.8)
```


## UMAP

Let us now vizualize our data by using UMAP.

```{r}
# install.packages("umap")
library(umap)
```

```{r,out.width="50%"}
umap=umap(data.pc)
plot(umap$layout,main="UMAP",col=colsamples,pch = 19,xlab = "UMAP1",ylab = "UMAP2")
legend(x="topright",legend=levels(as.factor(cell_types$CellType)),pch=19,col=brewer.pal(7, "Paired"),cex=0.8)
```

UMAP default parameters:
```{r,collapse=TRUE}
umap.defaults
```

For a description of each parameter type "help(umap.defaults)".

**Most relevant parameters**:

**n_neighbors** and **min_dist** are used to control the balance between local and global structure in the final projection.

**n_neighbors**: number of approximate nearest neighbors used to construct the initial high-dimensional graph. Low values will make UMAP to focus more on local structure.

**min_dist**: minimum distance between points in low-dimensional space. It controls how tightly UMAP clumps points together, with low values leading to more tightly packed embeddings and larger values making UMAP pack points together more loosely.

*Example: increase the min_dist parameter*

```{r,out.width="50%"}
umap=umap(data.pc,min_dist=0.5)
plot(umap$layout,main="UMAP",col=colsamples,pch = 19,xlab = "UMAP1",ylab = "UMAP2")
legend(x="topright",legend=levels(as.factor(cell_types$CellType)),pch=19,col=brewer.pal(7, "Paired"),cex=0.8)
```


