---
title: "Principal component regression - exercises"
output:
  html_document:
    df_print: paged
---

```{r,message=FALSE,warning=FALSE}
library(factoextra)
library(RColorBrewer)
```


### Principal component regression

Let us consider a dataset containing IC50 values (first column) for 232 cell lines (rows) that were treated with the drug Dabrafenib. IC50 values indicate the dose that is required to obtain a 50% reduction in cell viability, hence
the smaller the value the more effective is the drug. The other columns indicate the
expression of 18909 different genes in the various cell lines. 

```{r}
load(file="/Users/dsesia/Desktop/Teaching/Dimensionality reduction/Dabrafenib_gene_expression.RData")
y=data$y
X=data[,-1]
X[1:5,1:8]
```
Let us scale the data and perform PCA with the R function prcomp ("stats" package).
```{r}
X=scale(X)
pca=prcomp(X,scale = F) 
```

How much variance do the first PCs explain?
```{r,out.width="50%"}
fviz_eig(pca,ncp=10)
summ=summary(pca)
summ$importance[,1:20]
```

```{r}
data.plot=data.frame(pca$x)
colnames(data.plot)=c("PC1","PC2")

ggplot(data.plot, aes(x = PC1, y = PC2, colour =y)) +
  geom_point()+scale_colour_gradientn(colours = brewer.pal(11, "RdYlBu")) +theme_light()
```

Linear regression on the first 10 PCs. Notice that they are orthogonal!
```{r,out.width="50%"}
data.pcs=data.frame(cbind(y,pca$x[,1:10]))
heatmap(cor(data.pcs[,-1]),Rowv = NA,Colv = NA,main = "Correlation matrix",col=colorRampPalette(brewer.pal(8, "Blues"))(25))
legend(x="bottomright", legend=c("0", "1"),fill=colorRampPalette(brewer.pal(8, "Blues"))(2))
```

```{r}
lr=lm(y~.,data=data.pcs)
summary(lr)
```

Retrieve the coefficients for the original variables:
```{r}
coeff.pcr=as.matrix(lr$coefficients[-1])
beta <- pca$rotation[,1:10] %*% coeff.pcr
#temp=beta[order(abs(beta[,1]),decreasing = TRUE),,drop=FALSE]
```

*Exercise*: Compare (by dividing the dataset in a training and test set), the performance of PCR vs penalized regression.\

*Remark*: In general PCR will tend to do well in cases when the first few principal components are sufficient to capture most of the variation in the predictors.


